\Title

### 2023-02-10 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

In a recent review paper, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatio-temporal motifs within the neural activity may be useful representations to perform computations for various cognitive tasks. In 1982, Abeles asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the theory of rate coding, the second one emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronicity of activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, it's use for efficient coding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows for the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, but precise, relative spiking delays, i.e., the time between the arrival of an afferent spike at a given synapse and the contribution of the associated postsynaptic potential to the neuron's soma. Due to the variety of configurations and the possible coexistence of multiple superimposed motifs, this representation has a much higher information capacity than a firing rate-based approach to neural coding.

@@

To train and test our model, we need to define a dataset for which we control the direction and speed of motion. Let us now define a procedure for animating a natural scene with virtual eye movements, similar to those used in neurobiological~\citep{vinje_sparse_2000, baudot_animation_2013} and computational neuroscience~\citep{kremkow_push-pull_2016} studies. First, let's define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the visual field. In animals with a fovea, this is particularly useful because it allows the area with the highest density of photoreceptors in the environment to be moved, for example to a point of interest. A specific mechanism for this is called saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}). They are generated very quickly (about $80~\ms$) and have a wide range of speeds. On a more microscopic scale, the human gaze moves with a continuous drift similar to a Brownian trajectory~\citep{poletti_head-eye_2015}. In order to maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}.This approach first defines a finite set of possible 2D movements in polar coordinates. Based on the distribution of biological movements, we simplified it by selecting a set of eye movements as the Cartesian product of $8$ linearly spaced movement directions and 3 geometrically spaced speeds. Next, we define a gaze trajectory as segments whose duration is drawn from a Poisson distribution with a mean block length of $20~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, the trajectory is integrated, assuming first that the velocities are sampled uniformly and independently from the set of different motion sets, and second that the motion is uniform during a time segment. The resulting instances yield trajectories that are qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(left)).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} {\bf (Left)}~We use large natural images ($512\times512$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show an example path of the trajectory for $200$ time steps (green line). {\bf (Right)}~Snapshot of the synthetic event stream at different time steps. The dynamics of the sub-image as a function of time produces a naturalistic movie. This can be transformed into an event-based representation. Mimicking the retina, this representation encodes pixel-per-pixel proportional increases or decreases in luminance in the image, i.e. ON (red) and OFF (blue) events. In the lower left corner of the images, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the second and third image and that due to the aperture problems, contours parallel to the motion emit less spikes.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. % For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. 
For this purpose, we selected a database of 600 natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these have been pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $256 \times 256$ in size, and we will extract sub-images of size $64 \times 64$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 128~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space such that the subimage is translated using the position given by the trajectory at each time step without touching the borders. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity to correspond to a movement of one pixel per frame (i.e., per time bin).

To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(Right)). The goal here is to infer the correct motion solely by observing these events. 

>>>>

In a recent review paper, we reported experimental evidence for the presence of precise spiking motifs embedded in recordings from biological neural tissues~\cite{grimaldi_precise_2023}. These spatio-temporal motifs within the neural activity may be useful representations to perform computations for various cognitive tasks. In 1982, Abeles asked whether the role of cortical neurons is the integration of synaptic inputs or rather the detection of coincidences in temporal spiking motifs~\citep{abeles_role_1982}. While the first possibility favors the theory of rate coding, the second one emphasizes the function of temporal precision in the neural code. Since then, numerous studies have demonstrated the emergence of synchronicity of activity within a neuronal population~\citep{riehle_spike_1997, davis_spontaneous_2021}, efficient coding thanks to the use of spike latencies~\citep{perrinet_coding_2004, gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2002, carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and suggest the existence of spatio-temporal spiking motifs in biological spike trains. In neural models, the integration of heterogeneous delays~\citep{guise_bayesian_2014, zhang_supervised_2020, nadafian_bio-plausible_2020} allows for the efficient exploitation or detection of these spatio-temporal motifs embedded in the spike train. In particular,~\citet{izhikevich_polychronization_2006} introduced the notion of the polychronous group as a repetitive motif of spikes defined by a subset of neurons with different, but precise, relative spiking delays, i.e., the time between the arrival of an afferent spike at a given synapse and the contribution of the associated postsynaptic potential to the neuron's soma. Due to the variety of configurations and the possible coexistence of multiple superimposed motifs, representation with polychronous groups has a much higher information capacity than a firing rate-based approach to neural coding.

@@

To train and test our model, we need to define a dataset for which we control the direction and speed of motion. Let us now define a procedure for animating a natural scene with virtual eye movements, similar to those used in neurobiological~\citep{vinje_sparse_2000, baudot_animation_2013} and computational neuroscience~\citep{kremkow_push-pull_2016} studies. First, we draw a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the visual field. In animals with a fovea, this is particularly useful because it allows the area with the highest density of photoreceptors in the environment to be moved, for example to a point of interest. A specific mechanism for this is called saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}). They are generated very quickly (about $80~\ms$) and have a wide range of speeds. On a more microscopic scale, the human gaze moves with a continuous drift similar to a Brownian trajectory~\citep{poletti_head-eye_2015}. In order to maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}.This approach first defines a finite set of possible 2D movements in polar coordinates. Based on the distribution of biological movements, we simplified it by selecting a set of eye movements as the Cartesian product of $8$ linearly spaced movement directions and 3 geometrically spaced speeds. Next, we define a gaze trajectory as segments whose duration is drawn from a Poisson distribution with a mean block length of $20~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, the trajectory is integrated, assuming first that the velocities are sampled uniformly and independently from the set of different motion sets, and second that the motion is uniform during a time segment. The resulting instances yield trajectories that are qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(left)).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} {\bf (Left)}~We use large natural images ($512\times512$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show an example path of the trajectory for $200$ time steps (green line). {\bf (Right)}~Snapshots of the synthetic event stream at different time steps. The dynamics of the sub-image as a function of time produces a naturalistic movie. This can be transformed into an event-based representation. Mimicking the retina, this representation encodes pixel-per-pixel proportional increases or decreases in luminance in the image, i.e. ON (red) and OFF (blue) events. In the lower left corner of the snapshots, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the second and third image and that due to the aperture problems, contours parallel to the motion emit less spikes.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. % For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. 
For this purpose, we selected a database of 600 natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these have been pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $256 \times 256$ in size, and we will extract sub-images of size $64 \times 64$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 128~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space such that the subimage is translated using the position given by the trajectory at each time step without touching the borders. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity to correspond to a movement of one pixel per frame (i.e., per time bin).

To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. For a specific pixel and timestamp, an event is generated if the absolute value of this gradient exceeds a threshold. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(Right)). The goal here is to infer the correct motion solely by observing these events. 
